Modelling with ANIMO:\\ between Fuzzy Logic and Differential Equations

### 2015-06-29 Paper.tex
Changed the location of the figures to the end and removed them as required for submission to the journal

\pgfdeclareimage[height=3ex]{win-logo}{images/win_logo}
\pgfdeclareimage[height=3ex]{mac-logo}{images/mac_logo}
\pgfdeclareimage[height=3ex]{linux-logo}{images/linux_logo}
\pgfdeclareimage[height=3ex]{ubuntu-logo}{images/ubuntu_logo}
\pgfdeclareimage[height=2ex]{mac-command}{images/mac_command}
\pgfdeclareimage[height=2ex]{lower-bound}{images/lowerbound}
\pgfdeclareimage[height=2ex]{upper-bound}{images/upperbound_green}
\pgfdeclareimage[width=0.95\textwidth]{FigS10}{Figures/S10} %100TNF0EGF}{images/TNFa-EGF/100TNFa-0EGF/100TNF_0EGF_ANIMOvsFuzzy-crop}
\pgfdeclareimage[width=0.95\textwidth]{FigS11}{Figures/S11} %0TNF100EGF}{images/TNFa-EGF/0TNFa-100EGF/0TNF_100EGF_ANIMOvsFuzzy-crop}
\pgfdeclareimage[width=0.95\textwidth]{FigS12}{Figures/S12} %100TNF100EGF}{images/TNFa-EGF/100TNFa-100EGF/100TNF_100EGF_ANIMOvsFuzzy-crop}
\pgfdeclareimage[width=0.95\textwidth]{FigS6}{Figures/S6} %CircadianComparison}{images/Circadian/ANIMOvsODE-crop}
\pgfdeclareimage[width=0.9\textwidth]{Fig3}{Figures/3} %ANIMOvsODE-smaller-comparison}{images/Circadian/ANIMOvsODE_small_comparison-crop}

@@

\title{Modelling with ANIMO:\\ between Fuzzy Logic and Differential Equations}%ODE vs Fuzzy Logic: in medio stat ANIMO}%Formal models of biological networks in \animo2{}} 
%Building formal models of biological networks avoiding formal training}

@@

\author[
   addressref={aff1},                   % id's of addresses, e.g. {aff1,aff2}
   %corref={aff1},                       % id of corresponding address, if any
   noteref={n1},                        % id's of article notes, if any
   email={s.schivo@utwente.nl}   % email address
]{\inits{S}\fnm{Stefano}~\snm{Schivo}}
\author[
   addressref={aff2},
   noteref={n1},
   email={j.scholma@utwente.nl}
]{\inits{J}\fnm{Jetse}~\snm{Scholma}}
% \author[
%    addressref={aff3},
%    email={b.wanders@utwente.nl}
% ]{\inits{B}\fnm{Brend}~\snm{Wanders}}
% \author[
%    addressref={aff2},
%    email={rauc6788@hotmail.com}
% ]{\inits{RA}\fnm{Ricardo A.}~\snm{Urquidi Camacho}}
\author[
   addressref={aff3},
   email={p.e.vandervet@utwente.nl}
]{\inits{PE}\fnm{Paul E.}~\snm{van~der~Vet}}
\author[
   addressref={aff2},
   email={h.b.j.karperien@utwente.nl}
]{\inits{M}\fnm{Marcel}~\snm{Karperien}}
\author[
   addressref={aff2},
   %corref={aff2},
   email={j.n.post@utwente.nl}
]{\inits{JN}\fnm{Janine N.}~\snm{Post}}
\author[
   addressref={aff1},
   %corref={aff1},
   email={j.c.vandepol@utwente.nl}
]{\inits{J}\fnm{Jaco}~\snm{van~de~Pol}}
\author[
   addressref={aff1},
   corref={aff1},
   email={r.langerak@utwente.nl}
]{\inits{R}\fnm{Rom}~\snm{Langerak}}

@@

\begin{artnotes}
%\note{Sample of title note}     % note to the article
\note[id=n1]{Equal contributor} % note, connected to author
\end{artnotes}
%
% \let\oldthefootnote\thefootnote
% \renewcommand{\thefootnote}{\fnsymbol{footnote}}
% \footnotetext[1]{These authors contributed equally to this work.}
% \footnotetext[3]{To whom correspondence should be addressed.}
% \let\thefootnote\oldthefootnote

@@

\begin{abstract} % abstract <= 350 words. Headers: background, results, conclusions
\parttitle{Background}
Experimental biologists need to deal with large amounts of data on a daily basis.
This makes computational support essential in order
to reason on the dynamics of biological systems.
% The large amount of data an experimental biologist
% must deal with makes the use of computational methods essential
% to reason on the dynamics of complex systems.
The software tool ANIMO (Analysis of Networks with Interactive MOdelling)
provides such computational support and allows insight into the complex networks of signalling events occurring in living cells.
Biology experts are able to exploit the power of the underlying formal methods
via a user interface specifically tailored for biological applications.
\parttitle{Results}
ANIMO models show a degree of precision intermediate between ordinary differential equations (ODEs) and fuzzy logic.
We show this by presenting ANIMO models for two case studies: \emph{Drosophila melanogaster} circadian clock, and signal
transduction events downstream of TNF$\alpha$ and EGF in HT-29 human
colon carcinoma cells.
The models were originally developed with ODEs and
fuzzy logic respectively.
\parttitle{Conclusions}
% The software tool ANIMO, recently updated to work with the latest version of Cytoscape,
% allows the biologist to develop a formal model through a user friendly interface,
% achieving useful insight without the need to study any mathematical formalism.
%
ANIMO models replicate with good precision the results of both the ODE and
fuzzy logic models. Moreover, ANIMO models require less parameters than ODEs and are more precise than
fuzzy logic. For this reason we position the modelling paradigm of ANIMO
between ODEs and fuzzy logic.
\end{abstract}

@@

\section*{Background}\label{sec:introduction}
\subsection*{Modelling in cell biology}
Executable biology is a young subfield in computational modelling, aimed at constructing models that mimic
biological phenomena \emph{in silico}. It provides an interesting paradigm 
% that enriches the
% normally static biological network diagrams with dynamic information. An underlying formalism is used 
% to describe
% the dynamics of a network and its components. %, defining the executable model.
% The analysis of an executable model typically produces graphs and animations, which help understand how the network evolves with time.
% Executable models typically contain descriptions of the 
to enhance network diagrams with an underlying formal description of network components
and their interactions. Executable models typically contain descriptions of the 
direct interactions of network components, requiring the modeller to think in terms of
``cause and effect''. Biologists reason in similar ways about the molecular mechanisms of network
interactions, and this makes the construction of an executable model an intuitive process.
ANIMO (Analysis of Networks with Interactive MOdelling,~\cite{animo-ieee,animo-gene})
provides an enabling technology to increase the use of computational models by experimental
biologists using their domain-specific language. 
ANIMO enriches the normally static biological network diagrams with dynamic information,
which is then used to automatically produce formal models representing the biological network.
Such models are indispensable for formally
comparing experimental data with prior knowledge, or for structuring experimental findings into a
new theory.
When dealing with complex biological networks, executable biology models are particularly useful
to understand the non-linear dynamics and the entailed emergent properties of the networks.
In those cases, an ANIMO model can be used as a support to obtain insight based on abstract
representations of the interactions occurring inside living cells.
Other applications of ANIMO models include performing \emph{in silico} experiments and
the storage and transfer of knowledge on biological networks.

@@

\begin{figure*}[htbp]
\begin{center}
%    \includegraphics[width=.85\textwidth]{images/screenshotTNFmodelSmall}
  \includegraphics[width=.85\textwidth]{Figures/1}
\end{center}
\caption{\csentence{The Cytoscape user interface running the ANIMO plug-in.}
The \emph{Network} panel in the centre contains the nodes-edges
model of the example TNF$\alpha$ pathway (see Methods Section), with
colours indicating node activity levels and shapes representing different protein categories (see the \emph{Legend} on the left).
The \emph{Results Panel} on the right contains a graph plotting activity levels of selected nodes
during the first 24 hours of simulation of the model. The slider under the graph
allows the user to select the time instant (marked as a vertical red line in the graph) on which
the colours nodes in the \emph{Network} are based.
The series with the {\sf \_data} suffix is experimental
data from~\cite{pathway-compendium}, considering a treatment with 100~ng/ml TNF$\alpha$.\\
All acronyms used in this paper
and their corresponding UniProt IDs are listed in Suppl. Sect.~\ref{suppl-sec:names}.
\label{fig:cytoscape}}
\end{figure*}

@@

\begin{figure*}[htbp]
 \begin{center}
%   \includegraphics[width=0.9\textwidth]{images/modeling_methods_spectrum_animo}
  \includegraphics[width=0.9\textwidth]{Figures/2}
 \end{center}
\caption{\csentence{The spectrum of modelling methods, with the addition of ANIMO.}
The precision of ANIMO models is halfway between fuzzy logic and ODE. In addition, ANIMO
allows for an easier modelling experience thanks to a user-friendly interface based on
the widely used network modelling software Cytoscape.\label{fig:animo-spectrum}}
\end{figure*}

%
% In living cells, processes are regulated by networks of interacting molecules.
% Aberrations in these networks underlie a wide range of pathologies. The development of new therapies
% requires a thorough insight in the functioning of these networks. Obtaining such insight can be a challenging task.
% Feedback loops and crosstalk between pathways lead to an intricate wiring of the network.
% % Hence, it is necessary to study the ensemble of molecules involved, because the
% % behaviour of individual molecules is not sufficient for a complete understanding.
% Since the human brain is ill-suited to grasp the non-linear dynamics of these complex networks and
% the entailed emergent properties, the role of computational support is increasing
% in molecular biology.
% %
% %
% %
% % The systems biology approach to understanding biological systems starts off from a
% % scientific question and then follows an empirical cycle \--\ or rather a positive spiral \--\ of
% % knowledge/theory $\rightarrow$ model $\rightarrow$ hypotheses $\rightarrow$ experiments $\rightarrow$
% % observations $\rightarrow$ update and/or refinement of knowledge/theory,
% % until an answer to the original question is found (Figure~\ref{fig:empirical-spiral}).
% In this context, we can profit from models in different ways:
% \begin{enumerate}
%   \item to organize data and store knowledge,
%   \item to structure reasoning and discussion
%   \item to perform \emph{in silico} experiments and derive hypotheses.
% \end{enumerate}
% % An \emph{in silico} model is always a simplified representation of biological reality and is never the
% % aim in itself.
% % Rather, it is a powerful means in the process of gaining an understanding of the biological system.
% % Given its role in the empirical cycle, the process of modelling is especially effective
% % when applied by the experts with respect to a certain biological system. Biologists usually have a good sense
% % of cause-and-effect relationships of molecular interactions. In addition, they are the most knowledgeable
% % on the network topology and the dynamics of the biological system they are studying.
% % Since they also benefit most from the generation of hypotheses and from an efficient experimental design,
% % biologists would be the primary candidates to construct models of their research topic.
%
%
% As models are a formalization of knowledge or theories, an underlying formalism is needed to express
% this knowledge. Different formal methods have been successfully applied to construct representations
% of biological systems. Among these methods are Boolean logic~\cite{boolean-networks,boolean-networks-flower,boolean-networks2},
% ordinary differential equations (ODEs, reviewed in~\cite{hidde-review}),
% interacting state machines~\cite{interacting-sm1,interacting-sm2},
% process calculi~\cite{regev-shapiro,blenx,bio-pepa}, Timed Automata~\cite{ta-siebert,bartocci-oscillators,
% oded-ode-ta-discretization} and Petri nets~\cite{petri-nets0,petri-nets,petri-nets2}.
% Most of these formal methods have been implemented into software tools to aid the process
% of modelling. Due to the lack of such a supporting tool, \tas\ have remained a less
% frequently applied method.

% \subsection*{\tas\ for biological models}
% \tas\ have been developed to model the dynamic behaviour of systems with processes running in parallel~\cite{timed-automata-alur-dill}.
% As such, \tas\ have been applied in communication protocols and industrial control engineering~\cite{ta-audio-protocol,ta-wap-gateway,ta-ws-bap}. The
% parallels between these application areas and regulatory processes in cells have triggered the step towards
% their use in biology.
% In~\cite{ta-siebert}, the authors used \tas\ to extend a classical modelling paradigm~\cite{thomas-formalism},
% allowing to add temporal dynamics to gene network models.
% In~\cite{bartocci-oscillators}, a model of biological oscillators was described and synchronization properties
% were tested in this dynamic system.
% A discretization of ODEs to \tas\ was proposed in~\cite{oded-ode-ta-discretization}, applying
% a translation between the two formalisms to an example gene regulatory network. Two
% different approaches to transforming
% a Petri net model into \tas\ were presented in~\cite{ta-giapponesi},
% where the important issue of state space explosion was also addressed.
% Finally, the authors of~\cite{ta-radiazioni} proposed an \emph{ad hoc} \tas\ model of a radiation treatment
% system, which was then validated through the analysis tool UPPAAL~\cite{uppaal}.\\
% Each of these approaches has been successfully validated, demonstrating the potential of \tas\
% in biological applications. However, these approaches were all limited to simple
% or specific examples and none of these modelling methods
% has led to a tool implementation of the proposed method to encourage a broader use
% of \tas\ in molecular biology. We have recently introduced the software tool ANIMO
% (Analysis of Networks with Interactive MOdelling,~\cite{animo-ieee}), which aims
% at making available the power of \tas\ to the experimental biologists without requiring
% them to be fluent in any mathematical formalism.
%
% % Mastery of most existing modelling tools requires training and experience in mathematical modelling.
% % In this respect, a lack of tradition in quantitative
% % reasoning and formal methods within the biological community at large is still a stumbling block for
% % widespread application of modelling of biological systems. Here, we present an intuitive method for the
% % construction of formal in silico models of the dynamics of molecular networks, supported by a novel,
% % user friendly modelling tool, ANIMO (Analysis of Networks with Interactive
% % MOdelling,~\cite{animo-ieee}). \tas\ are used as the underlying mathematical formalism.
%
% % In the Methods section, we will explain how choosing a suitable abstraction level can make the construction
% % of models more intuitive. We will then show how ANIMO is designed to support the modelling process following
% % this approach. Construction of a small model based on experimental data will exemplify the method that we
% % propose. In the Results section, we first show an ANIMO model of the genes and proteins that constitute
% % the circadian clock network in \emph{Drosophila Melanogaster}. The remainder of that section is dedicated to
% % illustrate how a single modelling iteration in the empirical cycle is used to compile prior knowledge and
% % experimental data into a model, perform \emph{in silico} simulations and derive meaningful testable
% % hypotheses. These hypotheses are supported by literature on interactions in different cell types.
%
% In this paper we present \animo2{}, an updated version of ANIMO which allows for a considerably
% more efficient analysis of models. \animo2{} also introduces new features, such as the possibility to interrogate
% an ANIMO model with model checking queries represented as human language sentences instead of
% mathematical formulae. The case studies presented in this paper are centered around possible usage scenarios
% for ANIMO, and aim at illustrating how a biologist can profit from the tool in its latest iteration.
% An installation and user manual for \animo2{} is available at \url{http://fmt.cs.utwente.nl/tools/animo/manual.html}.

% \subsection*{A short introduction to ANIMO}
% {\Large [[briefly describe ANIMO, its modelling assumptions and its intended use. Refer to Gene paper?]]}

\section*{Results}\label{sec:results}
\subsection*{Modelling oscillation in Drosophila Melanogaster circadian clock}\label{sec:animo-drosophila}
To demonstrate that results obtained with ANIMO are comparable to results from widely used modelling
approaches, we present 
an ANIMO model of the circadian clock in \emph{Drosophila Melanogaster} (Suppl. Fig.~\ref{fig:drosophila-model}).
This ANIMO model is based on~\cite{drosophila-ode-model},
where ordinary differential equations (ODEs) were used.
The cyclic behaviour of the circadian clock is based on the alternating formation and destruction of the
CYC/CLK protein complex.
Concentration levels of this complex are in turn regulated by a series of proteins which are produced as
a consequence of CYC/CLK formation. The CWO protein
is central to the functioning of the network, as it degrades the mRNA for most of the involved proteins.
As such, CWO acts as an inhibitor that counterbalances the effect of CYC/CLK.
The positive influence of the light-regulated cryptochrome CRY on the degradation of TIM is a consequence
of the passage between day and night, allowing
the circadian clock to synchronize to a time zone (see Suppl. Sect.~\ref{suppl:repressilator}).

The output of the ANIMO model (see Suppl. Fig.~\ref{fig:drosophila-model}) matches the original ODE model.
In particular, the oscillations in both models show similar periods and phases (see Fig.~\ref{fig:grafici-drosophila-smaller} for
an example, and Suppl. Fig~\ref{suppl-fig:grafici-drosophila} for the complete comparison).
% Due to the compositional nature of \tas\, ANIMO allows for intuitive \emph{in silico} knock-out experiments,
% by right-clicking a node in the model and disabling it. Such experiments have been done
% before~\cite{drosophila-ode-model} and give similar results in our model.
Details of the comparison
between the ANIMO model and the original ODE model are given in Supplementary Section~\ref{suppl-sec:animo-drosophila}.
% \\
% {\Large [[Insert description of K.O. experiments with Drosophila model]]}

\begin{figure*}[!htpb]
\begin{center}
% \pgfuseimage{ANIMOvsODE-smaller-comparison}
  \pgfuseimage{Fig3}
\end{center}
\caption{\label{fig:grafici-drosophila-smaller}\csentence{Comparing the result of the ANIMO model of \emph{Drosophila Melanogaster} circadian clock
with the model of~\cite{drosophila-ode-model}.} 24 hours simulations of the two models were compared
against each other, synchronizing their start point as much as possible. The blue line is the ANIMO model
({\sf \_{}ANIMO} series), while the red line represents the data computed from the original ODE model
({\sf \_{}ODE} series). The vaules of all series were rescaled on a [0, 100] interval to facilitate comparison.}
\end{figure*}

@@

We constructed a model of the signalling network downstream of TNF$\alpha$ and EGF
in HT-29 human colon carcinoma cells, formalizing
the crosstalk that takes place between the pathways at different levels of cellular regulation.
We first modelled the two pathways in isolation~(Suppl. Figs.~\ref{fig:large-model-tnf}, \ref{fig:large-model-egf}),
using information on protein interactions from
the KEGG~\cite{kegg} and phosphosite~\cite{phosphosite} databases. These models were fitted to experimental data
from previous studies~\cite{pathway-compendium,pathway-autocrine}.
We then merged the two pathways into a single model and added autocrine crosstalk between the pathways that has been
described in~\cite{pathway-autocrine}.
Briefly, stimulation with TNF$\alpha$ ({\sf TNFa} in the model) leads to a rapid release of TGF$\alpha$ ({\sf TGFa}),
which activates the EGF receptor ({\sf EGFR}).
This activation causes secretion of IL-1$\alpha$ ({\sf IL-1a}) at later time points.
The effect of IL-1$\alpha$ is down-regulated by the secretion of IL-1 receptor antagonist ({\sf IL-1ra})
downstream of TNF$\alpha$.
The resulting model (Fig.~\ref{fig:large-model-all}A) was compared to the experimental data
for treatments with 100 ng/ml TNF alone and 100 ng/ml EGF alone (data not shown)~\cite{pathway-compendium}.

\def\largeModelScale{0.18}%0.15}%0.155}%0.27}
\def\legendScalaColori{0.21}%0.18}%0.21}
\def\legendScalaForme{0.21}%0.18}%0.21}
\def\scalaGrafici{0.0709}%0.13}
\begin{figure*}[!htpb]
\centering
%  \subfloat[\label{fig:large-model-no-hypotheses}]{\includegraphics[scale=\largeModelScale]{images/large_network_merged_no_hypotheses}}\\
% %\subfloat{\includegraphics[scale=\legendScalaColori]{images/legenda_forme}}\\ \subfloat{\includegraphics[scale=\legendScalaForme]{images/legenda_colori}}\\
% \subfloat{\includegraphics[scale=\legendScalaColori]{images/legenda_forme_e_colori}}\\
% \addtocounter{subfigure}{-1}
% \subfloat[\label{fig:large-model-graph1}]{\includegraphics[scale=\scalaGrafici]{images/TGF100_TNF0_ho_hyp_IL-1a3}}
% \subfloat[\label{fig:large-model-graph3}]{\includegraphics[scale=\scalaGrafici]{images/TNF5_C225_no_hyp_JNK1_MK2}}
%   \includegraphics[width=0.9\textwidth]{images/large_network_merged_no_hypotheses_completa}
  \includegraphics[width=0.9\textwidth]{Figures/4}
\caption{\csentence{Signalling network downstream of TNF$\alpha$ and EGF in human colon carcinoma cells.}
{\bf(A)} %{\bf \protect\subref{fig:large-model-no-hypotheses}}
The model for the merged TNF$\alpha$ and EGF pathways. Node colours represent the
activity level of the corresponding modelled reactants at time $t = 15$ minutes after
a stimulation of 100 ng/ml TNF$\alpha$ + 100 ng/ml EGF.
{\bf(B)} %{\bf \protect\subref{fig:large-model-graph1}}~
Modelled production of IL-1$\alpha$ after stimulation with 100 ng/ml TGF$\alpha$ (24 hours).
{\bf(C)} %{\bf \protect\subref{fig:large-model-graph3}}~
Modelled activation of JNK1 and MK2 after stimulation with 5 ng/ml TNF$\alpha$ + 10 $\mu$g/ml C225 (2 hours).
\\
The {\sf \_{}data} suffix identifies experimental data; all other series are computed by ANIMO.}\label{fig:large-model-all}
\end{figure*}

@@

Experimentally,
treatment with TGF$\alpha$ alone does not lead to secretion of IL-1$\alpha$. Instead, a co-stimulation with
TGF$\alpha$ and TNF$\alpha$ is required~\cite{pathway-autocrine}.
However, in the first version of the model, treatment with TGF$\alpha$ was sufficient for
IL-1$\alpha$ expression (Fig.~\ref{fig:large-model-all}B). Given the time delay until secretion of IL-1$\alpha$, it can be
expected that \emph{de novo} synthesis of IL-1$\alpha$ is required and that both
TNF$\alpha$ and TGF$\alpha$ are needed to activate transcription of the IL-1$\alpha$ gene.
JNK1 and ERK signal downstream of TNF$\alpha$ and TGF$\alpha$, respectively, and are known
to affect the activity of multiple transcription factors. We altered the model to make
activation of IL-1$\alpha$ expression dependent on both JNK1 activity and ERK activity
(Suppl. Fig.~\ref{fig:large-model-hypotheses}, arrows linking {\sf JNK1} and {\sf ERK} to {\sf IL-1a gene}).
After this modification to the model, IL-1$\alpha$ was no longer secreted
upon stimulation with TGF$\alpha$ alone, which greatly improved the fit between the measured IL-1$\alpha$
levels and the model (Fig.~\ref{fig:large-model-complete}B). This hypothesis could now be used to
design a new experiment to validate IL-1$\alpha$ as a target of combined JNK1 activity and ERK activity in
HT-29 cells. For example, kinase inhibitors specific to JNK1 and ERK could be used to confirm that activity of
both kinases is required for expression and secretion of IL-1$\alpha$. Performing the experiment is beyond
the scope of this study, but this hypothesis finds support in literature:
transcription factors c-Jun and c-Fos together
form a heterodimer known as AP-1 and are activated by JNK1 and ERK,
respectively~\cite{jnk-signaling,cfos-cjun}. AP-1 has been reported to bind to the
promoter of IL-1$\alpha$, providing evidence for a role in the regulation of IL-1$\alpha$
expression~\cite{ap1-il1a}. Based on these findings in literature we included c-Jun and
c-Fos in our model as transcriptional activators of IL-1$\alpha$ (Fig.\ref{fig:large-model-complete}A).

>>>>

\pgfdeclareimage[height=3ex]{win-logo}{images/win_logo}
\pgfdeclareimage[height=3ex]{mac-logo}{images/mac_logo}
\pgfdeclareimage[height=3ex]{linux-logo}{images/linux_logo}
\pgfdeclareimage[height=3ex]{ubuntu-logo}{images/ubuntu_logo}
\pgfdeclareimage[height=2ex]{mac-command}{images/mac_command}
\pgfdeclareimage[height=2ex]{lower-bound}{images/lowerbound}
\pgfdeclareimage[height=2ex]{upper-bound}{images/upperbound_green}
\pgfdeclareimage[width=0.95\textwidth]{FigS10}{Figures/S10}
\pgfdeclareimage[width=0.95\textwidth]{FigS11}{Figures/S11}
\pgfdeclareimage[width=0.95\textwidth]{FigS12}{Figures/S12}
\pgfdeclareimage[width=0.95\textwidth]{FigS6}{Figures/S6}

@@

\title{Modelling with ANIMO:\\ between Fuzzy Logic and Differential Equations}

@@

\author[
   addressref={aff1},                   % id's of addresses, e.g. {aff1,aff2}
   %corref={aff1},                       % id of corresponding address, if any
   noteref={n1},                        % id's of article notes, if any
   email={s.schivo@utwente.nl}   % email address
]{\inits{S}\fnm{Stefano}~\snm{Schivo}}
\author[
   addressref={aff2},
   noteref={n1},
   email={j.scholma@utwente.nl}
]{\inits{J}\fnm{Jetse}~\snm{Scholma}}
\author[
   addressref={aff3},
   email={p.e.vandervet@utwente.nl}
]{\inits{PE}\fnm{Paul E.}~\snm{van~der~Vet}}
\author[
   addressref={aff2},
   email={h.b.j.karperien@utwente.nl}
]{\inits{M}\fnm{Marcel}~\snm{Karperien}}
\author[
   addressref={aff2},
   %corref={aff2},
   email={j.n.post@utwente.nl}
]{\inits{JN}\fnm{Janine N.}~\snm{Post}}
\author[
   addressref={aff1},
   %corref={aff1},
   email={j.c.vandepol@utwente.nl}
]{\inits{J}\fnm{Jaco}~\snm{van~de~Pol}}
\author[
   addressref={aff1},
   corref={aff1},
   email={r.langerak@utwente.nl}
]{\inits{R}\fnm{Rom}~\snm{Langerak}}

@@

\begin{artnotes}
%\note{Sample of title note}     % note to the article
\note[id=n1]{Equal contributor} % note, connected to author
\end{artnotes}

@@

\begin{abstract} % abstract <= 350 words. Headers: background, results, conclusions
\parttitle{Background}
Experimental biologists need to deal with large amounts of data on a daily basis.
This makes computational support essential in order
to reason on the dynamics of biological systems.
The software tool ANIMO (Analysis of Networks with Interactive MOdelling)
provides such computational support and allows insight into the complex networks of signalling events occurring in living cells.
Biology experts are able to exploit the power of the underlying formal methods
via a user interface specifically tailored for biological applications.
\parttitle{Results}
ANIMO models show a degree of precision intermediate between ordinary differential equations (ODEs) and fuzzy logic.
We show this by presenting ANIMO models for two case studies: \emph{Drosophila melanogaster} circadian clock, and signal
transduction events downstream of TNF$\alpha$ and EGF in HT-29 human
colon carcinoma cells.
The models were originally developed with ODEs and
fuzzy logic respectively.
\parttitle{Conclusions}
ANIMO models replicate with good precision the results of both the ODE and
fuzzy logic models. Moreover, ANIMO models require less parameters than ODEs and are more precise than
fuzzy logic. For this reason we position the modelling paradigm of ANIMO
between ODEs and fuzzy logic.
\end{abstract}

@@

\section*{Background}\label{sec:introduction}
\subsection*{Modelling in cell biology}
Executable biology is a young subfield in computational modelling, aimed at constructing models that mimic
biological phenomena \emph{in silico}. It provides an interesting paradigm 
to enhance network diagrams with an underlying formal description of network components
and their interactions. Executable models typically contain descriptions of the 
direct interactions of network components, requiring the modeller to think in terms of
``cause and effect''. Biologists reason in similar ways about the molecular mechanisms of network
interactions, and this makes the construction of an executable model an intuitive process.
ANIMO (Analysis of Networks with Interactive MOdelling,~\cite{animo-ieee,animo-gene})
provides an enabling technology to increase the use of computational models by experimental
biologists using their domain-specific language. 
ANIMO enriches the normally static biological network diagrams with dynamic information,
which is then used to automatically produce formal models representing the biological network.
Such models are indispensable for formally
comparing experimental data with prior knowledge, or for structuring experimental findings into a
new theory.
When dealing with complex biological networks, executable biology models are particularly useful
to understand the non-linear dynamics and the entailed emergent properties of the networks.
In those cases, an ANIMO model can be used as a support to obtain insight based on abstract
representations of the interactions occurring inside living cells.
Other applications of ANIMO models include performing \emph{in silico} experiments and
the storage and transfer of knowledge on biological networks.

@@

@@

\section*{Results}\label{sec:results}
\subsection*{Modelling oscillation in Drosophila Melanogaster circadian clock}\label{sec:animo-drosophila}
To demonstrate that results obtained with ANIMO are comparable to results from widely used modelling
approaches, we present 
an ANIMO model of the circadian clock in \emph{Drosophila Melanogaster} (Suppl. Fig.~S4). %\ref{fig:drosophila-model}).
This ANIMO model is based on~\cite{drosophila-ode-model},
where ordinary differential equations (ODEs) were used.
The cyclic behaviour of the circadian clock is based on the alternating formation and destruction of the
CYC/CLK protein complex.
Concentration levels of this complex are in turn regulated by a series of proteins which are produced as
a consequence of CYC/CLK formation. The CWO protein
is central to the functioning of the network, as it degrades the mRNA for most of the involved proteins.
As such, CWO acts as an inhibitor that counterbalances the effect of CYC/CLK.
The positive influence of the light-regulated cryptochrome CRY on the degradation of TIM is a consequence
of the passage between day and night, allowing
the circadian clock to synchronize to a time zone (see Suppl. Sect.~D.2). %\ref{suppl:repressilator}).

The output of the ANIMO model (see Suppl. Fig.S4) %~\ref{fig:drosophila-model})
matches the original ODE model.
In particular, the oscillations in both models show similar periods and phases (see Fig.~\ref{fig:grafici-drosophila-smaller} for
an example, and Suppl. Fig~S6 %\ref{suppl-fig:grafici-drosophila}
for the complete comparison).
Details of the comparison
between the ANIMO model and the original ODE model are given in Supplementary Section D.3. %~\ref{suppl-sec:animo-drosophila}.

@@

We constructed a model of the signalling network downstream of TNF$\alpha$ and EGF
in HT-29 human colon carcinoma cells, formalizing
the crosstalk that takes place between the pathways at different levels of cellular regulation.
We first modelled the two pathways in isolation~(Suppl. Figs.~S7, S8), %\ref{fig:large-model-tnf}, \ref{fig:large-model-egf}),
using information on protein interactions from
the KEGG~\cite{kegg} and phosphosite~\cite{phosphosite} databases. These models were fitted to experimental data
from previous studies~\cite{pathway-compendium,pathway-autocrine}.
We then merged the two pathways into a single model and added autocrine crosstalk between the pathways that has been
described in~\cite{pathway-autocrine}.
Briefly, stimulation with TNF$\alpha$ ({\sf TNFa} in the model) leads to a rapid release of TGF$\alpha$ ({\sf TGFa}),
which activates the EGF receptor ({\sf EGFR}).
This activation causes secretion of IL-1$\alpha$ ({\sf IL-1a}) at later time points.
The effect of IL-1$\alpha$ is down-regulated by the secretion of IL-1 receptor antagonist ({\sf IL-1ra})
downstream of TNF$\alpha$.
The resulting model (Fig.~\ref{fig:large-model-all}A) was compared to the experimental data
for treatments with 100 ng/ml TNF alone and 100 ng/ml EGF alone (data not shown)~\cite{pathway-compendium}.

@@

Experimentally,
treatment with TGF$\alpha$ alone does not lead to secretion of IL-1$\alpha$. Instead, a co-stimulation with
TGF$\alpha$ and TNF$\alpha$ is required~\cite{pathway-autocrine}.
However, in the first version of the model, treatment with TGF$\alpha$ was sufficient for
IL-1$\alpha$ expression (Fig.~\ref{fig:large-model-all}B). Given the time delay until secretion of IL-1$\alpha$, it can be
expected that \emph{de novo} synthesis of IL-1$\alpha$ is required and that both
TNF$\alpha$ and TGF$\alpha$ are needed to activate transcription of the IL-1$\alpha$ gene.
JNK1 and ERK signal downstream of TNF$\alpha$ and TGF$\alpha$, respectively, and are known
to affect the activity of multiple transcription factors. We altered the model to make
activation of IL-1$\alpha$ expression dependent on both JNK1 activity and ERK activity
(Suppl. Fig.~S9, %\ref{fig:large-model-hypotheses}, 
arrows linking {\sf JNK1} and {\sf ERK} to {\sf IL-1a gene}).
After this modification to the model, IL-1$\alpha$ was no longer secreted
upon stimulation with TGF$\alpha$ alone, which greatly improved the fit between the measured IL-1$\alpha$
levels and the model (Fig.~\ref{fig:large-model-complete}B). This hypothesis could now be used to
design a new experiment to validate IL-1$\alpha$ as a target of combined JNK1 activity and ERK activity in
HT-29 cells. For example, kinase inhibitors specific to JNK1 and ERK could be used to confirm that activity of
both kinases is required for expression and secretion of IL-1$\alpha$. Performing the experiment is beyond
the scope of this study, but this hypothesis finds support in literature:
transcription factors c-Jun and c-Fos together
form a heterodimer known as AP-1 and are activated by JNK1 and ERK,
respectively~\cite{jnk-signaling,cfos-cjun}. AP-1 has been reported to bind to the
promoter of IL-1$\alpha$, providing evidence for a role in the regulation of IL-1$\alpha$
expression~\cite{ap1-il1a}. Based on these findings in literature we included c-Jun and
c-Fos in our model as transcriptional activators of IL-1$\alpha$ (Fig.\ref{fig:large-model-complete}A).

As a second example, we considered the behaviour of JNK1 and MK2. In the model, both
proteins were located downstream of TNF$\alpha$ but not TGF$\alpha$ or EGF. Hence, the
model did not show an effect of C225, a pharmacological inhibitor of ligand-EGFR
binding, on activation of JNK1 or MK2 after stimulation with TNF$\alpha$ (Fig.~\ref{fig:large-model-all}C). However, experimental
data show that C225 strongly reduces activation of JNK1 and MK2 upon stimulation with
TNF$\alpha$~\cite{pathway-autocrine}.
This fact is indicative of a role for EGFR in activation of JNK1 and MK2. Since both JNK1 and MK2
are located downstream of MEKK1, we hypothesized that activation
of MEKK1 is dependent on
both TNF$\alpha$-signalling and TGF$\alpha$-signalling. In the model we added a new
hypothetical node {\sf Hyp~2} (hypothesis~2) to link EGFR to MEKK1 (Suppl. Fig.~S9). %\ref{fig:large-model-hypotheses}).
This addition led to an improved fit of the model to the data upon treatment with TNF$\alpha$ + C225:
activation of both MK2 and JNK1 was strongly suppressed by C225 (Fig.~\ref{fig:large-model-complete}C).
Stimulation with EGF alone did not lead to activation of JNK1 and MK2.
These data support the validity of the modification to the model.
Further support for a link between EGFR and MEKK1 was found in literature. Specifically,
Ras has been reported as a direct activator of
MEKK1~\cite{ras-mekk1}. EGFR is a well-known and potent activator of Ras,
which is why it was already in our network~\cite{kegg}.
Other studies also report activation of JNK1 and phosphorylation of c-Jun downstream of Ras, which is consistent with
an interaction between Ras and MEKK1~\cite{cfos-cjun,ras-jnk1}.
Based on these findings, we adapted
our model by removing the {\sf Hyp~2} node and creating a direct interaction between Ras
and MEKK1 (Fig.~\ref{fig:large-model-complete}A). Experimentally, the role of Ras could be confirmed by using a
pharmacological inhibitor of Ras activity, and measuring the effect of this inhibitor on the activation of JNK1 and MK2.
Together, our model suggests that EGFR activity is required
but not sufficient for activation of JNK1 and MK2 in HT-29 cells.
